Relational Parametricity for Algebraically Indexed Types

### 2012-06-14 paper/section-general-framework.tex
Add program syntax for dealing with existential types, and extend the proof of the abstraction theorem
Small fixes
Macro-ize the names of the groups with dimensions, so we can change notation easily if need be

\begin{example}[Two-Dimensional Geometry]
  The two-dimensional geometry system has a sort for each of the
  geometric groups that we mentioned in
  \autoref{sec:motivating-examples}. Thus, $\SortSet = \{\SynTransl_2,
  \SynGL_2, \SynOrth_2, \SynGL_1 \}$. For the index operations, we
  have additive group structure on $\SynTransl_2$, multiplicative
  group structure on $\SynGL_1$, $\SynGL_2$ and $\SynOrth_2$,
  injections from $\SynOrth_2$ and $\SynGL_1$ into $\SynGL_2$,
  determinant and absolute value. Thus, $\IndexOpSet = \{ 0, +, -,
  1_G, -\cdot_G-, -^{-1_G}, \iota_O, \iota_1, \det, |\cdot| \}$, where
  $G \in \{\SynGL_1, \SynGL_2, \SynOrth_1\}$, with the following
  assignment of arities:
  \begin{displaymath}
    \begin{array}{@{}l@{\hspace{0em}=\hspace{0em}}l@{\hspace{0.5em}}l@{\hspace{0em}=\hspace{0em}}l}
      \indexOpArity(0) & ([], \SynTransl_2) &
      \indexOpArity(1_G) & ([], G) \\
      \indexOpArity(+) & ([\SynTransl_2, \SynTransl_2], \SynTransl_2) &
      \indexOpArity(\cdot_G) & ([G,G],G) \\
      \indexOpArity(-) & ([\SynTransl_2], \SynTransl_2) &
      \indexOpArity(^{-1_G}) & ([G], G) \\
      \indexOpArity(\iota_o) & ([\SynOrth_2], \SynGL_2) &
      \indexOpArity(\iota_1) & ([\SynGL_1], \SynGL_2) \\
      \indexOpArity(\det) & ([\SynGL_2], \SynGL_1) &
      \indexOpArity(|\cdot|) & ([\SynGL_1], \SynGL_1)
    \end{array}
  \end{displaymath}
  The intended interpretations of the top three pairs of operations
  are group unit, group combination and group negation,
  respectively. When we introduce equational theories for index
  expressions in \autoref{sec:type-equality} below, we will impose the
  (abelian) group laws. For this example, we have $\PrimTypeSet = \{
  \tyPrimNm{vec}, \tyPrimNm{real} \}$, with
  $\primTyArity(\tyPrimNm{vec}) = [\SynGL_2, \SynTransl_2]$ and
  $\primTyArity(\tyPrimNm{real}) = [\SynGL_1]$.
\end{example}

@@

\begin{example}[Two-Dimensional Geometry]
  Translations form an abelian group under addition, so we assume the
  abelian group axioms for translations:
  \begin{displaymath}
    \begin{array}{l}
      t : \SynTransl_2 \vdash t + 1 \stackrel{ax}\equiv t : \SynTransl_2 \\
      t_1, t_2, t_3 : \SynTransl_2 \vdash t_1 + (t_2 + t_3) \stackrel{ax}\equiv (t_1 + t_2) + t_3 : \SynTransl_2 \\
      t : \SynTransl_2 \vdash t + (-t) \stackrel{ax}\equiv 1 : \SynTransl_2 \\
      t_1, t_2 : \SynTransl_2 \vdash t_1 + t_2 \stackrel{ax}\equiv t_2 + t_1 : \SynTransl_2 \\
    \end{array}
  \end{displaymath}
  Similarly, the sort of scale factors $\SynGL_1$ forms an abelian
  group under multiplication, and the sorts $\SynGL_2$ and
  $\SynOrth_2$ form (non-abelian) multiplicative groups, so we assume
  the appropriate axioms. We also assume that the operations $\iota_O,
  \iota_1, \det$ and $|\cdot|$ are group homomorphisms, and that
  expressions of the form $\iota_1(s)$ commute with group
  multiplication in the sort $\SynGL_2$. The absolute value of the
  determinant of orthgonal transformations is always $1$, so we also
  assume $|\det(\iota_OO)| \stackrel{ax}\equiv 1$. Likewise, scaling
  maps have a determined determinant: $\det(\iota_1(s))
  \stackrel{ax}\equiv s \cdot s$.
\end{example}

@@

Given a relation environment $\rho \in \relEnv{E}(\Delta)$, we define
the set of extensions $\extends{\rho}{i\mathord:s}$ of $\rho$ by an
additional index variable $i\mathord:s$ to be
$\extends{\rho}{i\mathord:s} \isDefinedAs \{ \rho' \in
\relEnv{E}(\Delta,i\mathord:s) \sepbar \rho' \circ \pi^*_{i\mathord:s}
= \rho \}$. The set of extensions of a relational environment will be
used in the interpretation of the universal and existential type
formers.

@@

Note that both parts of \lemref{lem:tyeqsubst-relational} depend on
\lemref{lem:tyeqsubst-erasure} to be ``well-typed''.

@@

    \inferrule* [right=UnivApp]
    {\Delta; \Gamma \vdash M : \forall i\mathord:s. A \\ \Delta \vdash e : s}
    {\Delta; \Gamma \vdash M [e] : (\id_\Delta, e)^*A}
  \end{mathpar}

@@

We assign an index-erasure semantics to any well-indexed typing
context $\Delta \vdash \Gamma \isCtxt$ by induction on its structure:
$\ctxtSem{\epsilon} = \{*\}$ and $\ctxtSem{\Gamma, x : A} =
\ctxtSem{\Gamma} \times \tySem{A}$. For a well-typed program $\Delta;
\Gamma \vdash M : A$, we define the \emph{erasure interpretation} as a
function $\tmSem{e} : \ctxtSem{\Gamma} \to \tySem{A}$, completely
ignoring the indexing information. In light of
\lemref{lem:tyeqsubst-erasure}, we define this function directly on
the syntax of well-typed programs, rather than on the typing
derivations. We define $\tmSem{e}$ by the following clauses:
\begin{displaymath}
  \begin{array}{@{\hspace{0em}}l@{\hspace{0.5em}}c@{\hspace{0.5em}}l}
    \tmSem{x}\eta & \isDefinedAs & \eta_x \\
    \tmSem{(M,N)}\eta & \isDefinedAs & (\tmSem{M}\eta, \tmSem{N}\eta) \\
    \tmSem{\pi_1M}\eta & \isDefinedAs & \pi_1(\tmSem{M}\eta) \\
    \tmSem{\pi_2M}\eta & \isDefinedAs & \pi_2(\tmSem{M}\eta) \\
    \tmSem{\mathrm{inl}\ M}\eta & \isDefinedAs & \mathrm{inl}\ (\tmSem{M}\eta) \\
    \tmSem{\mathrm{inr}\ M}\eta & \isDefinedAs & \mathrm{inr}\ (\tmSem{M}\eta) \\
    \left\llbracket
      \begin{array}{l}
        \textrm{case}\ M\ \textrm{of}\\
        \textrm{inl}\ x.N_1;\\
        \textrm{inr}\ y.N_2
      \end{array}\right\rrbracket^{\mathit{tm}}\eta & \isDefinedAs &
    \left\{
      \begin{array}{ll}
        \tmSem{N_1}(\eta,a) & \textrm{if }\tmSem{M}\eta = \mathrm{inl}(a) \\
        \tmSem{N_2}(\eta,b) & \textrm{if }\tmSem{M}\eta = \mathrm{inr}(b)
      \end{array}
    \right. \\
    \tmSem{\lambda x. M}\eta & \isDefinedAs & \lambda v. \tmSem{M}(\eta, v) \\
    \tmSem{M N}\eta & \isDefinedAs & (\tmSem{M}\eta) (\tmSem{N}\eta) \\
    \tmSem{\Lambda i.\ M}\eta & \isDefinedAs & \tmSem{M}\eta \\
    \tmSem{M[e]}\eta & \isDefinedAs & \tmSem{M}\eta    
  \end{array}
\end{displaymath}
In the above, we have written $\eta_x$ to denote the appropriate
projection from $\eta$ to get the value in the environment
corresponding to the variable $x$ in the context.

>>>>

\begin{example}[Two-Dimensional Geometry]
  The two-dimensional geometry system has a sort for each of the
  geometric groups that we mentioned in
  \autoref{sec:motivating-examples}. Thus, $\SortSet = \{\SynTransl{2},
  \SynGL{2}, \SynOrth{2}, \SynGL{1} \}$. For the index operations, we
  have additive group structure on $\SynTransl{2}$, multiplicative
  group structure on $\SynGL{1}$, $\SynGL{2}$ and $\SynOrth{2}$,
  injections from $\SynOrth{2}$ and $\SynGL{1}$ into $\SynGL{2}$,
  determinant and absolute value. Thus, $\IndexOpSet = \{ 0, +, -,
  1_G, -\cdot_G-, -^{-1_G}, \iota_O, \iota_1, \det, |\cdot| \}$, where
  $G \in \{\SynGL{1}, \SynGL{2}, \SynOrth{2}\}$, with the following
  assignment of arities:
  \begin{displaymath}
    \begin{array}{@{}l@{\hspace{0em}=\hspace{0em}}l@{\hspace{0.5em}}l@{\hspace{0em}=\hspace{0em}}l}
      \indexOpArity(0) & ([], \SynTransl{2}) &
      \indexOpArity(1_G) & ([], G) \\
      \indexOpArity(+) & ([\SynTransl{2}, \SynTransl{2}], \SynTransl{2}) &
      \indexOpArity(\cdot_G) & ([G,G],G) \\
      \indexOpArity(-) & ([\SynTransl{2}], \SynTransl{2}) &
      \indexOpArity(^{-1_G}) & ([G], G) \\
      \indexOpArity(\iota_o) & ([\SynOrth{2}], \SynGL{2}) &
      \indexOpArity(\iota_1) & ([\SynGL{1}], \SynGL{2}) \\
      \indexOpArity(\det) & ([\SynGL{2}], \SynGL{1}) &
      \indexOpArity(|\cdot|) & ([\SynGL{1}], \SynGL{1})
    \end{array}
  \end{displaymath}
  The intended interpretations of the top three pairs of operations
  are group unit, group combination and group negation,
  respectively. When we introduce equational theories for index
  expressions in \autoref{sec:type-equality} below, we will impose the
  (abelian) group laws. For this example, we have $\PrimTypeSet = \{
  \tyPrimNm{vec}, \tyPrimNm{real} \}$, with
  $\primTyArity(\tyPrimNm{vec}) = [\SynGL{2}, \SynTransl{2}]$ and
  $\primTyArity(\tyPrimNm{real}) = [\SynGL{1}]$.
\end{example}

@@

\begin{example}[Two-Dimensional Geometry]
  Translations form an abelian group under addition, so we assume the
  abelian group axioms for translations:
  \begin{displaymath}
    \begin{array}{l}
      t : \SynTransl{2} \vdash t + 1 \stackrel{ax}\equiv t : \SynTransl{2} \\
      t_1, t_2, t_3 : \SynTransl{2} \vdash t_1 + (t_2 + t_3) \stackrel{ax}\equiv (t_1 + t_2) + t_3 : \SynTransl{2} \\
      t : \SynTransl{2} \vdash t + (-t) \stackrel{ax}\equiv 1 : \SynTransl{2} \\
      t_1, t_2 : \SynTransl{2} \vdash t_1 + t_2 \stackrel{ax}\equiv t_2 + t_1 : \SynTransl{2} \\
    \end{array}
  \end{displaymath}
  Similarly, the sort of scale factors $\SynGL{1}$ forms an abelian
  group under multiplication, and the sorts $\SynGL{2}$ and
  $\SynOrth{2}$ form (non-abelian) multiplicative groups, so we assume
  the appropriate axioms. We also assume that the operations $\iota_O,
  \iota_1, \det$ and $|\cdot|$ are group homomorphisms, and that
  expressions of the form $\iota_1(s)$ commute with group
  multiplication in the sort $\SynGL{2}$. The absolute value of the
  determinant of orthgonal transformations is always $1$, so we also
  assume $|\det(\iota_OO)| \stackrel{ax}\equiv 1$. Likewise, scaling
  maps have a determined determinant: $\det(\iota_1(s))
  \stackrel{ax}\equiv s \cdot s$.
\end{example}

@@

Given a relation environment $\rho \in \relEnv{E}(\Delta)$, we define
the set of extensions $\extends{\rho}{i\mathord:s}$ of $\rho$ by an
additional index variable $i\mathord:s$ to be
$\extends{\rho}{i\mathord:s} \isDefinedAs \{ \rho' \in
\relEnv{E}(\Delta,i\mathord:s) \sepbar \rho' \circ \pi_{i\mathord:s} =
\rho \}$. The set of extensions of a relational environment will be
used in the interpretation of the universal and existential type
formers.

@@

Note that both parts of \lemref{lem:tyeqsubst-relational} depend on
the corresponding parts of \lemref{lem:tyeqsubst-erasure} to be
``well-typed''.

@@

    \inferrule* [right=UnivApp]
    {\Delta; \Gamma \vdash M : \forall i\mathord:s. A \\ \Delta \vdash e : s}
    {\Delta; \Gamma \vdash M [e] : (\id_\Delta, e)^*A}

    \inferrule* [right=ExPack]
    {\Delta; \Gamma \vdash M : (\id_\Delta, e)^*A \\ \Delta, i\mathord:s \vdash A \isType}
    {\Delta; \Gamma \vdash \mathrm{pack}\ [e]\ M : \exists i\mathord:s. A}

    \inferrule* [right=ExUnpack]
    {\Delta; \Gamma \vdash M : \exists i\mathord:s. A \\
      \Delta, i\mathord:s; \pi_{i\mathord:s}^*\Gamma, x : A \vdash N : \pi_{i\mathord:s}^*B}
    {\Delta; \Gamma \vdash \mathrm{unpack}\ [i]\ x = M\ \mathrm{in}\ N : B}
  \end{mathpar}

@@

We assign an index-erasure semantics to any well-indexed typing
context $\Delta \vdash \Gamma \isCtxt$ by induction on its structure:
$\ctxtSem{\epsilon} = \{*\}$ and $\ctxtSem{\Gamma, x : A} =
\ctxtSem{\Gamma} \times \tySem{A}$. For a well-typed program $\Delta;
\Gamma \vdash M : A$, we define the \emph{erasure interpretation} as a
function $\tmSem{e} : \ctxtSem{\Gamma} \to \tySem{A}$, completely
ignoring the indexing information. In light of
\lemref{lem:tyeqsubst-erasure}, we define this function directly on
the syntax of well-typed programs, rather than on the typing
derivations. We define $\tmSem{e}$ by the following clauses:
\begin{displaymath}
  \begin{array}{@{\hspace{0em}}l@{\hspace{0.5em}}c@{\hspace{0.5em}}l}
    \tmSem{x}\eta & \isDefinedAs & \eta_x \\
    \tmSem{(M,N)}\eta & \isDefinedAs & (\tmSem{M}\eta, \tmSem{N}\eta) \\
    \tmSem{\pi_1M}\eta & \isDefinedAs & \pi_1(\tmSem{M}\eta) \\
    \tmSem{\pi_2M}\eta & \isDefinedAs & \pi_2(\tmSem{M}\eta) \\
    \tmSem{\mathrm{inl}\ M}\eta & \isDefinedAs & \mathrm{inl}\ (\tmSem{M}\eta) \\
    \tmSem{\mathrm{inr}\ M}\eta & \isDefinedAs & \mathrm{inr}\ (\tmSem{M}\eta) \\
    \left\llbracket
      \begin{array}{l}
        \textrm{case}\ M\ \textrm{of}\\
        \textrm{inl}\ x.N_1;\\
        \textrm{inr}\ y.N_2
      \end{array}\right\rrbracket^{\mathit{tm}}\eta & \isDefinedAs &
    \left\{
      \begin{array}{ll}
        \tmSem{N_1}(\eta,a) & \textrm{if }\tmSem{M}\eta = \mathrm{inl}(a) \\
        \tmSem{N_2}(\eta,b) & \textrm{if }\tmSem{M}\eta = \mathrm{inr}(b)
      \end{array}
    \right. \\
    \tmSem{\lambda x. M}\eta & \isDefinedAs & \lambda v. \tmSem{M}(\eta, v) \\
    \tmSem{M N}\eta & \isDefinedAs & (\tmSem{M}\eta) (\tmSem{N}\eta) \\
    \tmSem{\Lambda i.\ M}\eta & \isDefinedAs & \tmSem{M}\eta \\
    \tmSem{M[e]}\eta & \isDefinedAs & \tmSem{M}\eta \\
    \tmSem{\mathrm{pack}\ [e]\ M}\eta & \isDefinedAs & \tmSem{M}\eta \\
    \tmSem{\mathrm{unpack}\ [i]\ x = M\ \mathrm{in}\ N}\eta & \isDefinedAs & \tmSem{N}(\eta, \tmSem{M}\eta)
  \end{array}
\end{displaymath}
In the above, we have written $\eta_x$ to denote the appropriate
projection from $\eta$ to get the value in the environment
corresponding to the variable $x$ in the context.
